Thesis: draft

### 2021-05-26 draft.tex
latest update

\section{Real-world social networks}

@@

\chapter{Opinion dynamics}
\label{chap3}
The study of opinion dynamics in a group of people has two layers: a network layer and an opinion dynamics layer. The network layer, which describes the underlying structure and correlations between the actors, is described in the previous chapter, Chapter \ref{chap2}. This chapter focuses on the second layer, that of the opinion dynamics models. These models describe how opinions are formed and how they evolve in a group of people that interact socially. Hence, these opinion dynamics are the dynamics that run on the underlying network structure.\\
\newline
Several opinion dynamics models have been proposed; here we will review some of the most important ones. There are two groups of opinion dynamics models, namely binary or discrete models and continuous approaches. These will be discussed in Section \ref{Bin} and Section \ref{Con}, respectively. This thesis only deals with binary opinion dynamics models; the continuous models will be reviewed in order to give a more complete overview.\\
\newline
In section \ref{actMech}, we will pay attention to the activation mechanism. This is the mechanism that describes the timings in which information is exchanged between actors \cite{Perra2019}. Such mechanism changes the network from a static one to a temporal one, where actors become active and inactive over time. Then, in Section \ref{stubb}, the introduction of stubborn actors will be discussed. Section \ref{filter} will study filtering algorithms used by social media companies. This is needed because this thesis deals with opinion dynamics on on-line social networks, so the aspect of filtering cannot be ignored in the formation and evolution of opinions. Finally, in Section \ref{modelThesis} the opinion dynamics model and the filtering algorithms of this thesis will be summarized.

@@

One of the simplest opinion dynamics models is the binary majority model. Each actor has one of two opinions, A or B. Initially, these opinions can be equally distributed or not; the starting conditions can be balanced or unbalanced. Balanced starting conditions indicate the initial situation where $50 \%$ of the actors is given opinion A and $50 \%$ of the actors is given opinion B. The opinions are usually randomly assigned; this is however not necessary and opinions can e.g. also be assigned according to the community a certain actor belongs to. If the starting condition are unbalanced, we have a majority and a minority opinion. E.g. $80 \%$ of the actors can be assigned opinion A and $20 \%$ of the actors is then given opinion B. Hence, opinion A is the predominant or majority opinion and opinion B is the minority opinion. The following updating rule is imposed on the network: each actor adopts the majority opinion of its nearest neighbors \cite{Nguyen2020}. Initially each actor is assigned an opinion; then the opinions are evolved in time: each actor iteratively updates its opinion according to the majority opinion of its direct neighbors \cite{Nguyen2020}.\\
\newline
Several studies have been performed on this kind of model with the aim to find the conditions that drive the system to converge to unanimity or to see whether an initial majority opinion remains a majority opinion in the final state \cite{Nguyen2020}.\\
\newline
Most of these studies conclude that, in the stationary state, a consensus is obtained where all the actors adopt the same opinion \cite{Nguyen2020}. This is also the case when the opinions are equally and randomly distributed in an Erd\H{o}s-R\'{e}nyi network (with the side note that the degree needs to be high enough), as was shown by Benjamini et al. \cite{Benjamini2016}. It is also found that the rate of convergence increases with increasing edge density and that if the network is sparse enough, consensus will not be reached and the two opinions coexist \cite{Nguyen2020}. This can be explained by the fact that if the network is sparse enough, we may have disconnected parts; each of these disconnected groups may evolve to a different consensus opinion, so that over the entire network different opinions coexist \cite{Nguyen2020}.

@@

The probabilistic majority model is similar to the majority model, but has, as the name suggests, a probabilistic nature. In this model the fraction of neighbors of actor $i$ with opinion A, $p_\text{A}$, and the fraction of neighbors with opinion B, $p_\text{B}$, are determined. Then, the opinion of actor $i$ is updated: it takes on opinion A (or B) with a probability equal to $p_\text{A}$ (or $p_\text{B}$) \cite{Perra2019}.\\
\newline
For example, if $70\%$ of the neighbors of actor $i$ have an opinion A (and thus $30 \%$ of the neighbors have opinion B), actor $i$ will update its opinion to opinion A with a probability equal to $0.7$ and to opinion B with a probability equal to $0.3$ \cite{Perra2019}. Actor $i$ no longer updates its opinion according to the majority opinion carried by its neighbors, but the majority opinion still has a higher probability to be adopted by this actor.\\
\newline
This model is stochastic, which is an important component of opinion dynamics \cite{Castellano2009}\cite{Perra2019}. Opinion dynamics processes are not purely deterministic; individual behavior, social influence and influence of external factors such as mass media and political campaigns may vary in an unpredictable way \cite{Castellano2009}.

@@

In continuous opinion dynamics models, the opinion variable of the actors in the network is no longer discrete, in contrast to binary or discrete models, but takes real values (often in the interval $\interval{0}{1}$ or $\interval{-1}{1}$).\\
\newline
Some concepts that are valid for the discrete models no longer apply (e.g. the concept of a majority and minority opinion or equality of opinions); hence, they require a different framework \cite{Castellano2009}. \\
\newline
The basis of most continuous models is a set of $N$ actors, each having an opinion $x_i$ represented by a real number that takes values in some interval \cite{Castellano2009}. In contrast to discrete or binary models, all actors usually start with different opinions \cite{Castellano2009}. The possible outcomes are more complex than in the discrete models and often have opinion clusters emerging in the final states (where opinion clusters refer to clusters of actors with the same opinion) \cite{Castellano2009}. The number of opinion clusters can be one (consensus), two (polarization) or more (fragmentation) \cite{Castellano2009}.

@@

The model is based on a compromise strategy: after some discussion and a constructive debate, the opinions of the two actors move closer to each other \cite{Castellano2009}. The stationary state is a situation where several opinion clusters coexist. This is because the opinions move closer and closer to each other in several of such opinion clusters; once each opinion cluster is sufficiently far apart from the others (the difference in opinions between all actors in the two opinion clusters is larger than $\epsilon$), they no longer interact with each other \cite{Castellano2009}. Only actors inside the same opinion cluster keep on interacting with each other, which makes the opinions inside an opinion cluster converge to a single value \cite{Castellano2009}. Usually, the number and sizes of the opinion clusters depends on the threshold $\epsilon$, whereas $\mu$ determines the convergence rate \cite{Castellano2009}.\\

@@

The dynamics of this model is similar to the Deffuant model and the stationary state is also characterized by a number of opinion clusters \cite{Castellano2009}. The number of clusters decreases if $\epsilon$ increases and above a critical value $\epsilon_c$ of the threshold, there can only be one cluster \cite{Castellano2009}.\\

@@

Most of the standard opinion dynamics models use a static network in which actors are always active and exchanging opinions. In these models the opinion dynamics might go as follows: each time step a random actor is selected after which its opinion is updated according to the opinion dynamics model that is implemented. Another possibility is updating all actors at the same time during each time step. \\
\newline
It might be more realistic if actors become active/inactive over time. This would be a better representation of social media platforms, where users are active for some periods of time and then become inactive again. This kind of activation mechanism of the actors gives the network a temporal character. The underlying network structure does not change (each actor has a fixed number of neighbors that does not change over time; no new friends are made or, conversely, no old friends are being unfriended). However, by making actors active and inactive the exchange of information between actors becomes time dependent on which actors are active or not. Hence, this does affect the network structure in some way, because if an actor, for example, does not become active for a long period of time, it does not participate in the exchange of information and the network structure behaves like if this actor is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity and clustering may behave differently. \\
\newline
In this thesis, actors are made active with some probability $p_i$ every time step (thus, every time step, each actor becomes active with a probability $p_i$). The activation probability is set equal to $p = 0.1$, which is constant for all the actors. This constant activation probability is chosen for simplicity.

@@

In real life, some people are persistent towards their own opinion and are not open to the opinions of others. They are convinced that they carry the truth with them and that other people, with different world-views, are not capable of giving them any new and useful insights. They are close-minded in contrast to people with an open-mind that are willing to learn from others and to reconsider their beliefs. This is a bit simplistic and there are shades of gray between this black and white statement. However, there are groups in society, with more extreme opinions, that are know to be inflexible towards changing that (extreme) opinion (being stubborn and being extreme does not need to go hand in hand, you can be stubborn without being extreme and vice versa; however, it is often the case that there is some correlation between people with extreme opinions and being more persistent towards these opinions compared to people with more moderate beliefs; this idea of correlating extreme people with a stubborn character is often used in models \cite{Deffuant2006}\cite{Deffuant2002}\cite{Turner}) (having some resistance to change your opinion to any other opinion does not necessarily need to be a bad thing, it can be a form of critical thinking and a form of taking things with a grain of salt).\\ 
\newline
Knowing the impact of stubborn groups on the opinions of others is of importance, especially with the rise of social media. Social media facilitates the spread of information and ideas. This is, unfortunately, also true for extreme and persistent groups. They become capable of reaching and influencing larger groups of people. Before the era of social media, one could maybe argue that these extreme and stubborn groups were often isolated and therefore would not reach too many people. So their impact could only be minor (this is only a statement that could have been given, there are plenty of examples throughout human history that may contradict this argument; to only mention a few: Nazism in the 1930s and 1940s, the Ku Klux Klan in the late 19th and begin 20th century, communism exploited by, among other, Stalin and Mao in the 20th century, all examples of groups or regimes that can be considered extreme and that have had a (relatively) wide reach and a significant impact on society). Today, this argument is no longer valid; social media enables the contact with all sorts of beliefs and all sorts of people with almost no (time and spatial) constraints. Extreme and stubborn groups that may have been isolated before no longer need to be so and this may have unknown impact. \\
\newline
Several models that include some form of stubbornness have been proposed to study the impact of stubborn and extreme groups on society. These models aim to study whether the distribution of the stubborns (clustered versus random) matters; whether stubborn groups can lead to polarization or can overtake the opinions in the system or, contrary, whether stubbornness is protective against extremism and polarization. We will review the most important models.

@@

The bounded confidence models, discussed in Section \ref{bounded}, can be extended to included stubbornness. Guillaume Deffuant proposed such an extension where stubbornness is correlated to extremism \cite{Deffuant2006}. In his model, the actors have an opinion $x_i$ (which is a real number between $\interval{-1}{1}$) and a threshold $\epsilon_i$, which determines the range of opinions that can influence actor $i$ (actors with opinions between $\interval{x_i - \epsilon_i}{x_i + \epsilon_i}$ are capable of influencing actor $i$). He then introduced a fraction of extremists, $f_e$, which can have either the extreme opinion $-1$ or $+1$ \cite{Deffuant2006}. These extremists are given a threshold equal to $\epsilon_e$ \cite{Deffuant2006}. The other actors are considered to be moderate and can have opinions between $-1$ and $1$. The moderates are given a threshold $\epsilon$, with $\epsilon > \epsilon_e$ \cite{Deffuant2006}. This last condition implicitly denotes the resistance of the extremes towards other opinions in contrast to the more flexible nature of the moderates, which is based on the idea that people with extreme opinions will be more convinced of that opinion and will be less likely to change it.\\
\newline
Another example of introducing some kind of persistence towards the own opinion of the actors is the ``hardening of positions'' that was briefly discussed at the end of Section \ref{classMod}. The model introduced by Friedkin and Johnsen (Section \ref{classMod}) in which each actor adheres to its initial opinion to a certain degree can also be seen as introducing a form of stubbornness \cite{Noorazar2020}.

@@

Playing around with the threshold $\epsilon_i$ of the actors to include a notion of stubbornness, such as one can do for the bounded confidence models, is not possible for binary or discrete models. There are, however, other ways to include stubborn actors in these models. In the following, two different possibilities will be discussed. \\
\newline
In bounded confidence models, the link between stubbornness and extremism was often made and it was often assumed that the more extreme somebodies opinions are, the more persistent that somebody is towards those opinions. In binary models, however, there are only two possible opinions and in that sense it is less useful to talk about extreme and moderate opinions. However, we can still have people that rather keep their own opinion than be convinced by friends to switch to the other opinion.\\
\newline
\textbf{Resistance parameter}\\
\newline
One way of creating stubborn actors in the binary model is by giving each actor $i$ a resistance parameter $r_i$. This resistance parameter can have values between 0 and 1. If $r_i = 0$, the actor is not stubborn and can be easily convinced by the other opinion (according to the opinion dynamics model that is implemented: majority, probabilistic majority). If $r_i = 1$, the actor is completely stubborn and there is no chance of convincing it of the other idea. If $0 < r_i < 1$, the actor has some resistance but can still be convinced to switch to the other opinion, even though it will be less likely to do so. An uniform random number $f$ between 0 and 1 is drawn. If $f > r_i$, the actor will update its opinion; if $f \leqslant r_i$, the actor is not convinced and keeps its own opinion.\\
\newline
The resistance parameter $r_i$ can be different or the same for all actors. One can also control the fraction of stubborn actors, for example, there can be a fraction of actors that is completely stubborn ($r = 1$), while the rest is not stubborn ($r = 0$). Other possibilities can be found according to how the stubborn actors are distributed and how stubborn they are made. This makes the model flexible.\\
\newline
\textbf{The threshold model}\\
\newline
In the threshold model one introduces stubbornness by introducing a threshold parameter $T$. This parameter can be either the same for all the actors ($T$) or individualized ($T_i$). The main idea is that actors can only change their opinion to the other opinion if the fraction of friends with that opinion ($p$) is larger than $T$ ($p > T$). The idea behind the threshold model is that people are hesitant to change their opinion, but if they have enough friends with the other opinion, they might get convinced of that opinion as well. The more persistent you are, the more friends you need in order to be convinced of the other idea. \\
\newline
There are different possibilities of implementing this model. One way, which makes use of the majority model, is as follows. First the fraction of friends with opinion A and opinion B, $p_\text{A}$ and $p_\text{B}$ respectively, are determined; then the following rules are used to update the opinion: if $p_\text{A} > p_\text{B}$ and $p_\text{A} > T$, opinion A is adopted; if $p_\text{B} > p_\text{A}$ and $p_\text{B} > T$, opinion B is adopted; in all other cases, the current opinion of the actor is not changed.\\
\newline
If $T = 0$, this model reduces to the pure majority model described in Section \ref{majModel}. If $T = 1$, the actors do not change their opinion. For $0 < T < 1$, the model is a majority model, but the actors are more persistent towards their own opinion and need a larger fraction of friends to convince them to switch to the other viewpoint.

@@

We model opinion dynamics on social media platforms, with focus on Facebook. Each time we access Facebook we experience algorithmic personalization. The time-line or News Feed of each user on Facebook is the complex combination of posts from, among other, Facebook friends, companies and mass media. In 2013, Facebook reported that on average 1500 potential stories of friends and pages we follow or companies that are sponsored could be shown on the News Feed of each user \cite{facebook}. No user could ever process this amount of information, hence, Facebook selects around 300 of them, which then appear on your personalized News Feed. Their selection methods are based on (i) how often we interact with certain friends, pages or prominent figures; (ii) the amount of likes, shares and comments a post receives from your friends and others; (iii) how much you have interacted with certain types of posts before and (iv) whether you and other people on Facebook are hiding or reporting a certain post \cite{facebook}. \\
\newline
If we translate Facebook to a social network, we end up with a temporal network. Nodes represent the Facebook users, which become active and inactive over time. Edges represent friendship ties on Facebook, that is, they link an user to his Facebook friends and the pages he or she follows. Each node then has an attribute that needs to mimic the Facebook News Feed and which is susceptible to the filtering algorithm.

@@

	\Function{Update opinion}{$L_i$}
		\State \(\textsc{Order}(L_i)\) \Comment{Order hidden time-line according to REF, REC or PR filtering}
		\If{$\text{len}(L_i) > S$}
			\State \(R_i \leftarrow L_i.\textsc{resize}(S)\) \Comment{Time-line $R_i$ = $S$ first elements of ordered hidden time-line}
		\EndIf
		\State \(\text{op} \leftarrow \textsc{probMaj}(R_i)\) \Comment{Update opinion according to probabilistic majority model}
	\State \Return op \Comment{Return updated opinion}
	\EndFunction
\end{algorithmic}
\end{algorithm}
\noindent
Pseudo-code is given in Algorithm \ref{alg:update} to make the model more comprehensive and understandable. The pseudo-code for updating the opinions of the active actors in case of using a resistance parameter $r$ at time step $t$ is displayed.

@@

Real-world social networks often have high clustering and modular structure. We want to study the impact of these topologies on the dynamics of opinions. In this section, we will discuss the results related to different network topologies with focus on the effect of clustering and community structure. Two different opinion distributions are investigated, balanced and unbalanced starting conditions. For both starting conditions we investigate the evolution of the prevalence of the opinions with time and the formation of echo chambers. Two filtering algorithms will be compared, the REC and the PR filtering algorithms, in order to investigate the possible interplay between filtering algorithms and network structure.

@@

In this configuration $50 \%$ of the actors start with opinion A ($P_{\text{A}}(0) = 0.5$) and $50 \%$ of the actors start with opinion B ($P_{\text{B}}(0) = 0.5$).\\
\newline
\textbf{Evolution of the prevalence of the opinions}\\
\newline
We start with the results for the evolution of the prevalence of the opinions in case of an initial $50/50$ opinion distribution. For both the REC and the PR filtering algorithms, the community structure is not able to break the status quo for both of the filtering algorithms. The initial prevalence of the opinions is maintained over time (Fig. \ref{ev_op_50_50}). This is also the case for networks with high clustering \cite{Perra2019}.

>>>>

\section{Real-world social networks}\label{realWorld}

@@

\chapter{Opinion dynamics}
\label{chap3}
The study of opinion dynamics in a group of people has two layers: a network layer and an opinion dynamics layer. The network layer, which describes the underlying structure and correlations between individuals, is described in the previous chapter, Chapter \ref{chap2}. This chapter focuses on the second layer, that of the opinion dynamics models. These models describe how opinions are formed and how they evolve in a group of people that interact socially. Hence, these opinion dynamics are the dynamics that run on the underlying network structure.\\
\newline
Several opinion dynamics models have been proposed; here we will review some of the most important ones. There are two groups of opinion dynamics models, namely binary or discrete models and continuous approaches. These will be discussed in Section \ref{Bin} and Section \ref{Con}, respectively. This thesis only deals with binary opinion dynamics models; the continuous models will be reviewed in order to give a more complete overview.\\
\newline
In section \ref{actMech}, we will pay attention to the activation mechanism. This is the mechanism that describes the timings in which information is exchanged between actors \cite{Perra2019}. Such mechanism changes the network from a static to a temporal one, where actors become active and inactive over time. Then, in Section \ref{stubb}, the introduction of stubborn actors will be discussed. Section \ref{filter} will study filtering algorithms used by social media companies. This is needed because this thesis deals with opinion dynamics on on-line social networks, so the aspect of filtering cannot be ignored in the formation and evolution of opinions. Finally, in Section \ref{modelThesis} the opinion dynamics model and the filtering algorithms of this thesis will be summarized.

@@

One of the simplest opinion dynamics models is the binary majority model. Each actor has one of two opinions, A or B. Initially, these opinions can be equally distributed or not, i.e. the starting conditions can be balanced or unbalanced. Balanced starting conditions indicate the initial situation where $50 \%$ of the actors is given opinion A and $50 \%$ of the actors is given opinion B. The opinions are usually randomly assigned; this is however not necessary and opinions can e.g. also be assigned according to which community a certain actor belongs to. If the starting condition are unbalanced, we have a majority and a minority opinion. E.g. $80 \%$ of the actors can be assigned opinion A and $20 \%$ of the actors is then given opinion B. Hence, opinion A is the predominant or majority opinion and opinion B is the minority opinion. The following updating rule is imposed on the network: each actor adopts the majority opinion of its nearest neighbors \cite{Nguyen2020}. Initially each actor is assigned an opinion; then the opinions are evolved in time: each actor iteratively updates its opinion according to the majority opinion of its direct neighbors \cite{Nguyen2020}.\\
\newline
Several studies have been performed on this model with the aim to find the conditions that drive the system to converge to unanimity or to see whether an initial majority opinion remains a majority opinion in the final state \cite{Nguyen2020}.\\
\newline
Most of these studies conclude that, in the stationary state, a consensus is obtained where all the actors adopt the same opinion \cite{Nguyen2020}. This is also the case when the opinions are equally and randomly distributed in an Erd\H{o}s-R\'{e}nyi network (with the side note that the degree needs to be high enough), as was shown by Benjamini et al. \cite{Benjamini2016}. It is also found that the rate of convergence increases with increasing edge density and that if the network is sparse enough, consensus will not be reached and the two opinions coexist \cite{Nguyen2020}. This can be explained by the fact that if the network is sparse enough, we may have disconnected parts; each of these disconnected groups may evolve to a different consensus opinion, so that over the entire network different opinions coexist \cite{Nguyen2020}.

@@

The probabilistic majority model is similar to the majority model, but has, as the name suggests, a probabilistic nature. In this model the fraction of neighbors of actor $i$ with opinion A, $p_\text{A}$, and the fraction of neighbors with opinion B, $p_\text{B}$, are determined. Then, the opinion of actor $i$ is updated: it takes on opinion A (or B) with a probability equal to $p_\text{A}$ (or $p_\text{B}$) \cite{Perra2019}.\\
\newline
For example, if $80\%$ of the neighbors of actor $i$ have an opinion A (and thus $20 \%$ of the neighbors have opinion B), actor $i$ will update its opinion to opinion A with a probability equal to $0.8$ and to opinion B with a probability equal to $0.2$ \cite{Perra2019}. Actor $i$ no longer updates its opinion according to the majority opinion carried by its neighbors, but the majority opinion still has a higher probability to be adopted by this actor.\\
\newline
This model is stochastic, which is an important component of opinion dynamics \cite{Castellano2009}\cite{Perra2019}. Opinion dynamics processes are not purely deterministic; individual behavior, social influence and influence of external factors such as mass media and political campaigns may vary in an unpredictable way \cite{Castellano2009}.

@@

In continuous opinion dynamics models, the opinion variable of the actors in the network is no longer discrete, in contrast to binary or discrete models, but takes real values (often in the interval $\interval{0}{1}$ or $\interval{-1}{1}$).\\
\newline
Some concepts that are valid for discrete models no longer apply (e.g. the concept of a majority and minority opinion or equality of opinions); hence, they require a different framework \cite{Castellano2009}. \\
\newline
The basis of most continuous models is a set of $N$ actors, each having an opinion $x_i$ represented by a real number that takes values in some interval \cite{Castellano2009}. In contrast to discrete or binary models, all actors usually start with different opinions \cite{Castellano2009}. The possible outcomes are more complex than in the discrete models and often have opinion clusters emerging in the final states (where opinion clusters refer to clusters of actors with the same opinion) \cite{Castellano2009}. The number of opinion clusters can be one (consensus), two (polarization) or more (fragmentation) \cite{Castellano2009}.

@@

The model is based on a compromise strategy: after some discussion and a constructive debate, the opinions of the two actors move closer to each other \cite{Castellano2009}. The stationary state is a situation where several opinion clusters coexist. The opinions move closer and closer to each other in several of such opinion clusters; once each opinion cluster is sufficiently far apart from the others (the difference in opinions between all actors in the two opinion clusters is larger than $\epsilon$), they no longer interact with each other \cite{Castellano2009}. Only actors inside the same opinion cluster keep on interacting with each other, which makes the opinions inside an opinion cluster converge to a single value \cite{Castellano2009}. Usually, the number and sizes of the opinion clusters depends on the threshold $\epsilon$, whereas $\mu$ determines the convergence rate \cite{Castellano2009}.\\

@@

The dynamics is similar to the Deffuant model and the stationary state is also characterized by a number of opinion clusters \cite{Castellano2009}. The number of clusters decreases if $\epsilon$ increases and above a critical value $\epsilon_c$ of the threshold, there can only be one cluster \cite{Castellano2009}.\\

@@

Most of the standard opinion dynamics models use a static network in which actors are always active and exchanging opinions. In these models the opinion dynamics might go as follows: each time step a random actor is selected after which its opinion is updated according to the opinion dynamics model that is implemented. Another possibility is updating all actors at the same time during each time step. \\
\newline
It might be more realistic if actors become active/inactive over time. This would be a better representation of social media platforms, where users are active for some periods of time and then become inactive again. This activation mechanism of the actors gives the network a temporal character. The underlying network structure does not change (each actor has a fixed number of neighbors that does not change over time; no new friends are made or, conversely, no old friends are being unfriended). However, by making actors active and inactive the exchange of information between actors becomes time dependent on which actors are active or not. Hence, this does affect the network structure in some way, because if an actor, for example, does not become active for a long period of time, it does not participate in the exchange of information and the network structure behaves like if this actor is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity and clustering may behave differently. \\
\newline
In this thesis, actors are made active with some probability $p_i$ every time step (thus, every time step, each actor becomes active with a probability $p_i$). The activation probability is set equal to $p = 0.1$, which is constant for all the actors. This constant activation probability is chosen for simplicity.

@@

In real life, some people are persistent towards their own opinion and are not open to the opinions of others. They are convinced that they carry the truth with them and that other people, with different world-views, are not capable of giving them any new and useful insights. They are close-minded in contrast to people with an open-mind that are willing to learn from others and to reconsider their beliefs. This is a bit simplistic and there are shades of gray between this black and white statement. However, there are groups in society, with more extreme opinions, that are know to be inflexible towards changing that (extreme) opinion (being stubborn and being extreme does not need to go hand in hand, you can be stubborn without being extreme and vice versa; however, it is often the case that there is some correlation between people with extreme opinions and being more persistent towards these opinions compared to people with more moderate beliefs; this idea of correlating extreme people with a stubborn character is often used in models \cite{Deffuant2006}\cite{Deffuant2002}\cite{Turner}) (having some resistance to change your opinion to another opinion does not necessarily need to be a bad thing, it can be a form of critical thinking and taking things with a grain of salt).\\ 
\newline
Knowing the impact of stubborn groups on the opinions of others is of importance, especially with the rise of social media. Social media facilitates the spread of information and ideas. This is, unfortunately, also true for persistent groups with extreme ideas: they become capable of reaching and influencing larger groups of people. Before the era of social media, one could maybe argue that these extreme and stubborn groups were often isolated and therefore would not reach too many people. So their impact could only be minor (this is only a statement that could have been given, there are plenty of examples throughout human history that may contradict this argument; to only mention a few: Nazism in the 1930s and 1940s, the Ku Klux Klan in the late 19th and begin 20th century, communism exploited by, among other, Stalin and Mao in the 20th century, all examples of groups or regimes that can be considered extreme and that have had a (relatively) wide reach and a significant impact on society). Today, this argument is no longer valid; social media enables the contact with all sorts of beliefs and all sorts of people with almost no (time and spatial) constraints. Extreme and stubborn groups that may have been isolated before no longer need to be so and this may have unknown impact. \\
\newline
Several models that include stubbornness have been proposed to study the impact of stubborn and extreme groups on society. These models aim to study whether the distribution of the stubborns (clustered versus random) matters; whether stubborn groups can lead to polarization or can overtake the opinions in the system or, contrary, whether stubbornness is protective against extremism and polarization. We will review the most important models.

@@

The bounded confidence models, discussed in Section \ref{bounded}, can be extended to include stubbornness. Guillaume Deffuant proposed such an extension where stubbornness is correlated to extremism \cite{Deffuant2006}. In his model, the actors have an opinion $x_i$ (which is a real number between $\interval{-1}{1}$) and a threshold $\epsilon_i$, which determines the range of opinions that can influence actor $i$ (actors with opinions between $\interval{x_i - \epsilon_i}{x_i + \epsilon_i}$ are capable of influencing actor $i$). He then introduced a fraction of extremists, $f_e$, which can have either the extreme opinion $-1$ or $+1$ \cite{Deffuant2006}. These extremists are given a threshold equal to $\epsilon_e$ \cite{Deffuant2006}. The other actors are considered to be moderate and can have opinions between $-1$ and $1$. The moderates are given a threshold $\epsilon$, with $\epsilon > \epsilon_e$ \cite{Deffuant2006}. This last condition implicitly denotes the resistance of the extremes towards other opinions in contrast to the more flexible nature of the moderates, which is based on the idea that people with extreme opinions will be more convinced of that opinion and will be less likely to change it.\\
\newline
Another example of introducing persistence towards the own opinion of the actors is the ``hardening of positions'' that was briefly discussed at the end of Section \ref{classMod}. The model introduced by Friedkin and Johnsen (Section \ref{classMod}) in which each actor adheres to its initial opinion to a certain degree can also be seen as introducing stubbornness \cite{Noorazar2020}.

@@

Playing around with the threshold $\epsilon_i$ of the actors to include stubbornness, such as one can do for the bounded confidence models, is not possible for binary or discrete models. There are, however, other ways to include stubborn actors in these models. In the following, two different possibilities will be discussed. \\
\newline
In bounded confidence models, the link between stubbornness and extremism was often made and it was often assumed that the more extreme somebodies opinions are, the more persistent that somebody is towards those opinions. In binary models there are only two possible opinions and in that sense it is less useful to talk about extreme and moderate opinions. However, we can still have people that rather keep their own opinion than be convinced by friends to switch to the other opinion.\\
\newline
\textbf{Resistance parameter}\\
\newline
One way of creating stubborn actors in the binary model is by giving each actor $i$ a resistance parameter $r_i$. This resistance parameter can have values between 0 and 1. If $r_i = 0$, the actor is not stubborn and can be easily convinced by the other opinion. If $r_i = 1$, the actor is completely stubborn and there is no chance of convincing it of the other idea. If $0 < r_i < 1$, the actor has some resistance but can still be convinced to switch to the other opinion, even though it will be less likely to do so. An uniform random number $f$ between 0 and 1 is drawn. If $f > r_i$, the actor will update its opinion; if $f \leqslant r_i$, the actor is not convinced and keeps its own opinion.\\
\newline
The resistance parameter $r_i$ can be different or the same for all actors. One can also control the fraction of stubborn actors, for example, there can be a fraction of actors that is completely stubborn ($r = 1$), while the rest is not stubborn ($r = 0$). Other possibilities can be found according to how the stubborn actors are distributed and how stubborn they are made. This makes the model flexible.\\
\newline
\textbf{The threshold model}\\
\newline
In the threshold model one introduces stubbornness by introducing a threshold parameter $T$. This parameter can be either the same for all the actors ($T$) or individualized ($T_i$). The main idea is that actors can only change their opinion to the other opinion if the fraction of friends with that opinion ($p$) is larger than $T$ ($p > T$). The idea behind the threshold model is that people are hesitant to change their opinion, but if they have enough friends with the other opinion, they might get convinced of that opinion as well. The more persistent you are, the more friends you need in order to be convinced of the other idea. \\
\newline
There are different possibilities of implementing this model. One way, which makes use of the majority model, is as follows. First the fraction of friends with opinion A and opinion B, $p_\text{A}$ and $p_\text{B}$ respectively, are determined; then the following rules are used to update the opinion: if $p_\text{A} > p_\text{B}$ and $p_\text{A} > T$, opinion A is adopted; if $p_\text{B} > p_\text{A}$ and $p_\text{B} > T$, opinion B is adopted; in all other cases, the current opinion of the actor is not changed.\\
\newline
If $T = 0$, this model reduces to the pure majority model described in Section \ref{majModel}. If $T = 1$, the actors do not change their opinion. For $0 < T < 1$, the model is a majority model, but the actors are more persistent towards their own opinion and need a larger fraction of friends to convince them to switch to the other viewpoint.

@@

We model opinion dynamics on social media platforms, with focus on Facebook. Each time we access Facebook we experience algorithmic personalization. The time-line or News Feed of each user on Facebook is the complex combination of posts from, among other, Facebook friends, companies and mass media. In 2013, Facebook reported that on average 1500 potential stories of friends and pages we follow or companies that are sponsored could be shown on the News Feed of each user \cite{facebook}. No user could ever process this amount of information, hence, Facebook selects around 300 of them, which then appear on your personalized News Feed. Their selection methods are based on (i) how often we interact with certain friends, pages or prominent figures; (ii) the amount of likes, shares and comments a post receives from our friends and others; (iii) how much we have interacted with certain types of posts before and (iv) whether we and other people on Facebook are hiding or reporting a certain post \cite{facebook}. \\
\newline
If we translate Facebook to a social network, we end up with a temporal network. Nodes represent the Facebook users, which become active and inactive over time. Undirected edges represent friendship ties on Facebook, that is, they link an user to his Facebook friends and the pages he or she follows. Each node then has an attribute that needs to mimic the Facebook News Feed and which is susceptible to the filtering algorithm.

@@

	\Function{Update opinion}{$L_i$}
		\State \(\textsc{Order}(L_i)\) \Comment{Order hidden time-line according to REF, REC or PR filtering}
		\If{$\text{len}(L_i) > S$}
			\State \(R_i \leftarrow L_i.\textsc{resize}(S)\) \Comment{Time-line $R_i$ = $S$ first elements of ordered hidden time-line}
		\EndIf
		\State \(\text{op} \leftarrow \textsc{probMaj}(R_i)\) \Comment{Update opinion according to probabilistic majority model}
	\State \Return op \Comment{Return updated opinion}
	\EndFunction
\end{algorithmic}
\end{algorithm}
\noindent
The pseudo-code for updating the opinions of the active actors in case of using a resistance parameter $r$ at time step $t$ is displayed in Algorithm \ref{alg:update} to make the model more comprehensive and understandable.

@@

Real-world social networks often have high clustering and modular structure. We want to study the impact of these topologies on the dynamics of opinions. In this section, we will discuss the results related to different network topologies with focus on the effect of clustering and community structure. Two different opinion distributions are investigated, balanced and unbalanced starting conditions. For both starting conditions we investigate the evolution of the prevalence of the opinions and the formation of echo chambers. Two filtering algorithms will be compared, the REC and the PR filtering algorithms, in order to investigate the interplay between filtering algorithms and network structure.

@@

In this configuration $50 \%$ of the actors start with opinion A ($P_{\text{A}}(0) = 0.5$) and $50 \%$ of the actors start with opinion B ($P_{\text{B}}(0) = 0.5$).\\
\newline
\textbf{Evolution of the prevalence of the opinions}\\
\newline
We start with the results for the evolution of the prevalence of the opinions. For both the REC and the PR filtering algorithms, the community structure is not able to break the status quo. The initial prevalence of the opinions is maintained over time (Fig. \ref{ev_op_50_50}). This is also the case for networks with high clustering \cite{Perra2019}.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_10x100_01-0001_003-0008_50-50_PR_REC_8x7.png}
    \label{sbm_50-50}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10x100_10-001-0001_50-50_PR_REC.png}
    \label{sbm-ws_50-50}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) for balanced starting conditions.]{Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) for balanced starting conditions. \textbf{(a)} SBM; \textbf{(b)} SBM-WS. Two configurations are implemented for the SBM: a high modularity ($Q = 0.908 \pm 0.005$)) and a low modularity ($Q = 0.220 \pm 0.007$) configuration. For each model the REC and the PR filtering algorithms are compared. The dotted black line represents the initial prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_50_50}
\end{figure}
\noindent
\textbf{Echo chambers}\\
\newline
Fig. \ref{echo_50_50} shows the distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$. The region on the $x$-axis close to zero describes neighborhoods in which none, or few actors, adopt opinion B (these are thus neighborhoods in which most actors adopt opinion A). Conversely, the region on the $x$-axis close to one describes neighborhoods in which a majority of the actors adopt opinion B. The distribution is obtained by imposing ten equispaced bins, $\Delta x = 0.1$ \cite{Perra2019}.
